---
title: "Brazilian municipalities road distances matrix"
bibliography: references.bib
---

## Introduction

Distance information between places is useful to evaluate the proximity and interconnection of regions. The Euclidean distance between two places, although simple and easy to compute, is not realistic in terms of dislocation costs. This dataset present the **road distance metrics** between all Brazilian municipalities.

## Methods

As reference to starting and ending point of the routes, I opted to use the municipality seat localization ("sede dos municípios") as reference, which is usually the city hall or historic downtown localization. This data is originally generated by the Brazilian Institute of Geography and Statistics (IBGE) and available on the [{geobr}](https://ipeagit.github.io/geobr/) package. The last available update is for the year of 2010[^1] and present 5,565 features.

[^1]: As new data for the 2022 Brazilian Census be available, the dataset will be likely updated.

With the list of municipalities and respective seat coordinates, a list of pairs of municipalities as computed using simple combinatorial analysis. Example:

```{r}
x <- c("a", "b", "c", "d")

combn(x, m = 2)
```

::: callout-note
With this, I assume that the distance between two municipalities is the same, independently of the direction.
:::

Thus, for 5,565 municipalities, we have a total of $5,565 \times 5,565 / 2 = 15,484,612$ routes to be computed.

```{r}
choose(n = 5565, k = 2)
```

To compute the road distance between pairs of municipalities, the [OSRM](http://project-osrm.org/) API service was used, with the [{osrm}](https://cran.r-project.org/package=osrm) package [@osrm_paper_2022]. More specifically, the table service was used, considering the "car" profile, returning as result the distance in **meters** and estimated trip duration in **minutes** for the shortest direct route possible.

### Route example

For example, lets compute the road distance between Rio de Janeiro, RJ and São Paulo, SP with the {osrm} package.

```{r}
#| message: false
#| warning: false
library(geobr)
library(osrm)
library(leaflet)
library(sf)

seats <- read_municipal_seat(showProgress = FALSE)

rj <- subset(seats, code_muni == 3304557)
sp <- subset(seats, code_muni == 3550308)

route <- osrmRoute(
  src = rj, dst = sp, 
  overview = "full",
  osrm.profile = "car"
)
```

```{r}
# Route distance, in meters
route$distance
```

```{r}
# Route duration, in minutes
route$duration
```

```{r}
route |>
  st_transform(4326) |>
  leaflet() |>
  addTiles() |>
  addPolylines()
```

::: callout-important
For some pairs of municipalities, the OSRM service was not able to determine a possible road route. This is expected, as some municipalities are reachable only by plane or boat.
:::

The scripts used to prepare the dataset is available [here](https://github.com/rfsaldanha/distbrasil).

## Dataset

The dataset with the road distances and trip duration are available on Zenodo. The current version was computed between May and June 2024.

You can download the dataset directly from Zenodo or download it directly from R, using the [{zendown}](https://rfsaldanha.github.io/zendown/) package.

```{r}
library(zendown)

# zen_file() |> readRDS()
```

## Future plans

-   Compute routes using other available routing services.

-   Yearly updates of the dataset, as the road infrastructure may change.
