---
title: "Dengue outbreak in Brazil, 2024: understanding the numbers"
date: "2024-02-05"
categories: [epidemiology, dengue]
---

I am writing this post to take a look at the dengue cases in Brazil, as the news reports that there is an increasing number of *suspected cases* early on 2024.

## Packages

```{r}
#| message: false
library(tidyverse)
library(microdatasus)
```

## Data

Let's download the most recent data from the Brazilian Health Ministry with the [microdatasus](https://rfsaldanha.github.io/microdatasus/) package. **Premiminary datasets** are available for the years of 2023 and 2024.

```{r}
dengue_raw <- fetch_datasus(
  year_start = 2023, year_end = 2024, 
  information_system = "SINAN-DENGUE", 
  timeout = 480
)

# vars = c("DT_NOTIFIC", "DT_SIN_PRI", "MUNICIPIO", "CLASSI_FIN")

dengue <- process_sinan_dengue(
  data = dengue_raw, 
  municipality_data = FALSE
)
```

## Overview

Let's take a look at the suspect cases.

```{r}
dengue |>
  mutate(
    DT_SIN_PRI = as_date(DT_SIN_PRI),
    DT_SIN_PRI = ceiling_date(DT_SIN_PRI, unit = "month")
  ) |>
  filter(year(DT_SIN_PRI) >= 2023) |>
  group_by(DT_SIN_PRI) |>
  summarise(freq = n()) |>
  ungroup() |>
  ggplot(aes(x = DT_SIN_PRI, y = freq)) +
  geom_bar(stat = "identity")
```

```{r}
res <- dengue |>
  filter(year(DT_SIN_PRI) >= 2024) |>
  select(DT_SIN_PRI, CLASSI_FIN) |>
  mutate(CLASSI_FIN = replace_na(CLASSI_FIN, "Em investigação"))
```

```{r}
res %>%
  group_by(CLASSI_FIN) %>%
  summarise(freq = n())
```
